---
title: "USMortality Project Data Analysis"
author: "Wenxiao Zhou"
date: "4/19/2021"
output: word_document
---

Load the Data From 2015-2019
```{r}
setwd("/Users/zhouwenxiao/Desktop/USMortality")
load("USall.rData")
load("US19.rData")
```

Overall Analysis

(1)Age-adjusted Mortality Rate for Year 2015-2019 among overall group and category by sex
<point out why choosing age-adjusted rate: weighted on age group make more sense to balance difference mortality counts within groups
https://www.health.ny.gov/diseases/chronic/ageadj.htm>

<Age-adjusted data source: (add reference) https://data.cdc.gov/NCHS/NCHS-Death-rates-and-life-expectancy-at-birth/w9j2-ggv5/data>

To see the time trend more clearly, we include the data from 1999-2018 under race categories, sex categories for age-adjusted mortality rate.

```{r}
library(ggplot2)
library(tidyverse)
overall<-read.csv("age_mortality.csv")
overall$Mortality<-as.numeric(overall$Mortality)
all_race<-overall %>% subset(Race=="All Races") 
colnames(all_race)<-c("Year","Race","Sex","Life","Mortality_Rate")
white<-overall %>% subset(Race=="White")
colnames(white)<-c("Year","Race","Sex","Life","Mortality_Rate")
black<-overall %>% subset(Race=="Black")
colnames(black)<-c("Year","Race","Sex","Life","Mortality_Rate")
```

```{r}
#sex comparison overall race level
p1<-ggplot(all_race,aes(Year,Mortality_Rate,color=Sex)) +
  geom_line() +
  geom_point(size=3,shape=21) +
  xlab('Year')+ylab('Mortality Age-adjusted Rate (deaths per 100,000)') +
  labs(list(color="Sex")) +
  theme_minimal() +
  ggtitle('Age-adjusted Mortality Rate(Sex Comparison) in 1999-2018')
p1
```
```{r}
#race & sex comparison overall years
race<-rbind(white,black)
race$category<-ifelse(race$Race=="White" & race$Sex=="Male",'W-M',
                  ifelse(race$Race=="White" & race$Sex=="Female",'W-F',
                  ifelse(race$Race=="Black" & race$Sex=="Male",'B-M','B-F')))
race$category<-as.factor(race$category)
p2<-ggplot(race,aes(Year,Mortality_Rate,color=category)) +
   geom_line() +
  geom_point(size=3,shape=21) +
  xlab('Year') + ylab('Mortality Age-adjusted Rate (deaths per 100,000)') +
  labs(list(color="Categories")) +
  theme_minimal()+
  ggtitle('Age-adjusted Mortality Rate(Sex,Race Comparison) in 1999-2018')
p2
```



(2) Frequency Tables among marital status, resident status, education factors
There is no significant difference among years under the same level of the factor. But for same year, the level differences are significant. <description by texts,summary>
```{r}
#Basic frequency tables of marital, resident, education factors
USall$Year<-as.factor(USall$Year)
t1<-prop.table(table(USall$Martial,USall$Year),2)
t2<-prop.table(table(USall$Resident,USall$Year),2)
t3<-prop.table(table(USall$Education,USall$Year),2)
```

Here, we find the distribution of mortality counts within education levels are worth further dicussed. So, we focus on education factors, categories as four new categories: high_school and below, bachelor or college, master or PHD, unknown. And draw the Stacked Bar plot to see the trend clearly.
<http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html>

Enrollment Data: https://data.census.gov/cedsci/table?q=school&tid=ACSST1Y2019.S1401

```{r}
library(viridis)
library(hrbrthemes)
overall_edu<-USall
teen<-which(overall_edu$Education %in% c("0","11"))
high<-which(overall_edu$Education %in% c("21","31"))
college<-which(overall_edu$Education %in% c("41","51","61"))
grad<-which(overall_edu$Education %in% c("71","81"))
unknown<-which(overall_edu$Education=="91")
overall_edu$Education<-ifelse(overall_edu$ID %in% teen,"teen",
          ifelse(overall_edu$ID %in% high, "high",
          ifelse(overall_edu$ID %in% college,"college",
          ifelse(overall_edu$ID %in% grad, "grad","unknown"))))
overall_edu1<-overall_edu[which(overall_edu$Education!="unknown"),]
edu<-data.frame(table(overall_edu1$Education,overall_edu1$Year))
edu$enroll<-c(41874863,17086446,18620217,4036762,
              41857908,17108087,18436671,4169611,
              41828973,17150113,18105312,4188939,
              41747569,17100878,17792902,4275426,
              41703677,16932635,17507427,4321881)
edu$pro<-edu$Freq/edu$enroll
edu$pro<-round(edu$pro,digits=4)
colnames(edu)<-c("Education","Year","Frequency","Enroll","Prop")
ggplot(edu,aes(fill=Education,y=Prop,x=Year)) +
  geom_bar(position="stack",stat="identity") +
  scale_fill_viridis(discrete=T) +
  ggtitle("Education Level Detection") +
  theme_ipsum() +
  xlab("")
```

(3)Time Tendency of Mortality Causes
```{r}
#Adults Suicide
timea<-USall
timea$Race<-ifelse(timea$Race=="1","White",
                    ifelse(timea$Race=="2","Others","Black")) 
timea$Age<-as.integer(timea$Age)
timea<-subset(timea,Age>=30 & Age<=37,
              select=c(Sex,Age,Cause,Race,Month,Year))
timea2<-data.frame(table(timea$Cause,timea$Month,timea$Year))
colnames(timea2)<-c("Cause","Month","Year","Freq")
timea2<-timea2[order(timea2$Year,timea2$Freq,decreasing=TRUE),]
timea1<-timea %>% filter(Cause=="40")
ti1<-data.frame(table(timea1$Month,timea1$Year))
colnames(ti1)<-c("Month","Year","Freq")
bartlett.test(Freq~Month,data=ti1)
comp1<-aov(Freq~Month, data = ti1)
summary(comp1)
#multiple comparison
library(asbio)
pairw.anova(y=ti1$Freq,x=ti1$Month,method="lsd",MSE=8742,df.err=48)
```

```{r}
#Adults Heart Disease
timea3<-timea %>% filter(Cause=="21")
ti3<-data.frame(table(timea3$Month,timea3$Year))
colnames(ti3)<-c("Month","Year","Freq")
bartlett.test(Freq~Month,data=ti3)
comp3<-aov(Freq~Month, data = ti3)
summary(comp3)
#multiple comparison
library(asbio)
pairw.anova(y=ti3$Freq,x=ti3$Month,method="lsd",MSE=29888,df.err=48)
```

(4) Further discussion and conclusion in overall analysis see on the data report.




Year 2019 Analysis<空间维度的分析，涉及病因>

(1-1) Top mortality causes of adults:
use '39 Causes of Death
First of all, define adults as age between 20-59 years, includes the boundaries.

```{r}
US19$Age<-as.integer(US19$Age)
US19$Race<-ifelse(US19$Race=="1","White",
                    ifelse(US19$Race=="2","Others","Black"))
adult<-subset(US19,Age>=30 & Age<=37,
              select=c(Sex,Age,Cause,Race,Month))

adult1<-data.frame(table(adult$Cause,adult$Sex))
colnames(adult1)<-c("Cause","Sex","Freq")
adult1<-adult1[order(adult1$Sex,adult1$Freq,
                       decreasing=TRUE),]
#adult top 5 causes: Male
#39(All other and unspecified accidents and adverse effects)
#37(All other diseases)
#21(Ischemic heart diseases)
#40(Intentional self-harm (suicide))
#15(Other malignant neoplasms)

#Female
#37(All other diseases)
#39(All other and unspecified accidents and adverse effects)
#15(Other malignant neoplasms)
#21(Ischemic heart diseases)
#09(Malignant neoplasm of breast)

#For adults, what if we do not consider the effects of Sex
#overall top 5 causes
adult2<-data.frame(table(adult$Cause))
adult2<-adult2[order(adult2$Freq,decreasing=TRUE),]
#39(All other and unspecified accidents and adverse effects)
#37(All other diseases)
#21(Ischemic heart diseases)
#40(Intentional self-harm (suicide))
#15(Other malignant neoplasms)
```



(1-2) Top mortality causes of olders:
use '39 Causes of Death
First of all, define adults as age between 60+ years, includes the boundaries.

```{r}
old<-subset(US19,Age>=38,
              select=c(Sex,Age,Cause,Race,Month))
old1<-data.frame(table(old$Cause,old$Sex))
colnames(old1)<-c("Cause","Sex","Freq")
old1<-old1[order(old1$Sex,old1$Freq,
                       decreasing=TRUE),]
#adult top 5 causes: Male
#37(All other diseases)
#21(Ischemic heart diseases)
#22(Other diseases of heart)
#15(Other malignant neoplasms)
#28(Chronic lower respiratory diseases)

#Female
#37(All other diseases)
#21(Ischemic heart diseases)
#22(Other diseases of heart)
#17(Alzheimer’s disease)
#24(Cerebrovascular diseases)

#For elderly, what if we do not consider the effects of Sex
#overall top 5 causes
old2<-data.frame(table(old$Cause))
old2<-old2[order(old2$Freq,decreasing=TRUE),]
#37(All other diseases)
#21(Ischemic heart diseases)
#22(Other diseases of heart)
#28(Chronic lower respiratory diseases)
#15(Other malignant neoplasms)
```



(2)State Level Mortality Analysis

data source: https://www.cdc.gov/nchs/nvss/vsrr/provisional-tables.htm

US 2019 population:<mortality rate calculation>
https://www.census.gov/search-results.html?searchType=web&cssp=SERP&q=population%20by%20states

```{r}
library(readxl)
library(plotly)
library(Hmisc)
state2019<-read_excel("state2019.xlsx")
state<-select(state2019,State,Total,Population)
state$State<-capitalize(tolower(state$State))
state$Rate<-round(state$Total/state$Population,digits=4)
state$hover <- with(state, 
               paste(State, '<br>', "Mortality Rate",Rate,
               '<br>',"Deaths",Total))
state$code<-c("AL","AK","AZ","AR","CA","CO","CT","DE","DC","FL","GA","HI","ID","IL","IN","IA","KS","KY","LA","ME","MD","MA","MI","MN","MS","MO","MT","NE","NV","NH","NJ","NM","NY","NC","ND","OH","OK","OR","PA","RI","SC","SD","TN","TX","UT","VT","VA","WA","WV","WI","WY")
# give state boundaries a white border
l <- list(color = toRGB("white"), width = 2)
# specify some map projection/options
g <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('white')
)

fig <- plot_geo(state, locationmode ='USA-states')
fig <- fig %>% add_trace(
    z = ~Rate, text = ~hover, locations = ~code,
    color = ~Rate, colors = 'Purples'
  )
fig <- fig %>% colorbar(title = "Mortality Rates")
fig <- fig %>% layout(
    title = '2019 US Motality Rate Levels by States',
    geo = g
  )
fig
```
save as a web page for interactively observe on the plot



(3)Recall the overall analysis on education levels and mortality rates, we discuss that teen and high are two groups with highest mortality rates:
```{r}
edu2019<-overall_edu1 %>% filter(Year=="2019")
edu2019$Race<-ifelse(edu2019$Race=="1","White",
                    ifelse(edu2019$Race=="2","Others","Black"))
teen2019<-edu2019 %>% filter(Education=="teen")
high2019<-edu2019 %>% filter(Education=="high")
table(teen2019$Manner)
table(high2019$Manner)
```
From two tables studies, Accidents and Natural Deaths are two main categories of status, we only focus on these two categories.

```{r}
teendeath<-teen2019 %>% filter(Manner %in% c("7","1"))
teena<-data.frame(table(teendeath$Cause,teendeath$Race))
colnames(teena)<-c("Cause","Race","Freq")
teena<-teena[order(teena$Race,teena$Freq,
                       decreasing=TRUE),]
#top 5 causes: 
#White
#37(All other diseases)
#21(Ischemic heart diseases)
#22(Other diseases of heart)
#28(Chronic lower respiratory diseases)
#24(Cerebrovascular diseases)

#Others
#37,21,24,22,15(Other malignant neoplasms)

#Black
#37,21,33(Certain conditions originating in the perinatal period),22,24

highdeath<-high2019 %>% filter(Manner %in% c("7","1"))
higha<-data.frame(table(highdeath$Cause,highdeath$Race))
colnames(higha)<-c("Cause","Race","Freq")
higha<-higha[order(higha$Race,higha$Freq,
                       decreasing=TRUE),]
#top 5 causes: 
#White
#37,21,22,28,39(All other and unspecified accidents and adverse effects)

#Others
#37,21,39,22,15

#Black
#37,21,22,24,39
                
```

From overall years comparison as well as 2019 analysis, Heart Disease always be the top 3 mortality causes among different education levels, different race and age groups. This disease can be represented the whole trend of mortality rates among states.

We draw the heart disease mortality by state:

Data source:
https://www.cdc.gov/nchs/pressroom/sosmap/heart_disease_mortality/heart_disease.htm

```{r}
library(maps)
heart2019<-read.csv("/Users/zhouwenxiao/Desktop/USMortality/heart2019.csv")
statesMap<-map_data("state")
heart2019$STATE<-tolower(heart2019$STATE)
heartMap<-inner_join(statesMap,heart2019,by=c("region"="STATE"))
ggplot(heartMap, aes(x = long, y = lat, group = group, fill = RATE)) + geom_polygon(color = "black") + ggtitle("Heart Disease Mortality Rate by States in 2019")+
  scale_fill_gradient(low = "white", high = "red", guide = "legend")
```

Make comparison of education levels by states:

<Reference: 
The impact of increasing education levels on rising life expectancy https://genus.springeropen.com/articles/10.1186/s41118-019-0055-0>

Education rank Data Resource: https://wallethub.com/edu/e/most-educated-states/31075>

Basing on the article,we detect whether there is relationship between states' education levels and mortality rates：
```{r}
educate2019<-read_excel("/Users/zhouwenxiao/Desktop/USMortality/educate2019.xlsx")
statesMap<-map_data("state")
educate2019$State<-tolower(educate2019$State)
eduMap<-inner_join(statesMap,educate2019,by=c("region"="State"))
names(eduMap)<-c("long","lat","group","order","region","subregion",
                 "rank","score","educational","quality")
ggplot(eduMap, aes(x = long, y = lat, group = group)) + 
  geom_polygon(fill = "white", color = "black")
ggplot(eduMap, aes(x = long, y = lat, group = group, fill = score)) + geom_polygon(color = "black")+ggtitle("Education Evaluation by States")+
  scale_fill_gradient(low = "blue", high = "white", guide = "legend")
```

From the plot, state with high mortality rate in heart disease has low education score. Relationship between these two factors verify.





